Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-14 paper/paper.tex
The lemmas seem to work. Finally.

\begin{definition}[Equivalent path graphs]
Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $\abs{X} \le k$.
\end{definition}

@@

Pruned de~Bruijn graphs can be built by merging nodes of a de~Bruijn graph. They are similar to manifold de Bruijn graphs \cite{Lin2014}.

\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
\end{lemma}

\begin{proof}
Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $\abs{X} \le k$, let $G'' = (V'', E'')$ be the path graph after the changes in the key set, and let $v''_{K} \in V''$ be the node with $G''.\gkey(v''_{K}) = K$.

Each node $v \in V' \setminus V'_{K}$ is also a node of graph $G''$ with $G''.\gvalue(v) = G'.\gvalue(v)$. As each occurrence of substring $K$ in the \kcollection{k} $\mathcal{S}$ can be extended into an occurrence of substring $G'.\gkey(v)$ for a node $v \in V'_{K}$, we also have $G''.\gvalue(v''_{K}) = G'.\gvalue(v)$ for all $v \in V'_{K}$.

We now prove that for every path matching the pattern in graph $G'$, we can find a matching path in graph $G''$ with the same set of occurrences, and the other way around. As graph $G'$ is a pruned de~Bruijn graph and we prove that graph $G''$ is \kequivalent{k} to it, graph $G''$ must also be a pruned de~Bruijn graph.

(From $G'$ to $G''$) Consider a path $P' = v'_{0} \dotsm v'_{\abs{X}-1}$ matching the pattern in graph $G'$. Let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be the sequence of nodes of graph $G''$ created by replacing every node $v \in V'_{K}$ in path $P$ with node $v''_{K}$. We note that $G''.\gkey(v''_{i})$ is a prefix of $G'.\gkey(v'_{i})$ for $0 \le i < \abs{X}$, and hence $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$ is a prefix of $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$ for $0 \le i < \abs{X}-1$. As key $G'.\gkey(v'_{i})$ prefix-matches $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$, key $G''.\gkey(v''_{i})$ must also prefix-match $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$, giving us an edge $(v''_{i}, v''_{i+1}) \in E''$. Hence sequence $P''$ is a path in $G''$, with $G''.\glabel(P'') = G'.\glabel(P')$ and $G''.\gvalue(v''_{0}) = G'.\gvalue(v'_{0})$.

% FIXME the same for P'' -> P'
% by contradiction: assume that G'' contains X but G' does not
% let X[i, ...] be the rightmost suffix not in G'
% X[i, ...] does not occur in the k-collection but X[i+1, ...] does
% X[i] G''.key(i+1) does occur, must be a proper prefix of X[i, ...]
% ...
(From $G''$ to $G'$)
\end{proof}

% Pruning generalizes the graph in the same way as DBG generalizes it?

\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P$ be a path in $G'$ with $\abs{P'} \le k$. Then there is a path $P$ in graph $G$ with $G.\glabel(P) = G'.\glabel(P')$.
\end{lemma}

% FIXME proof
\begin{proof}
Foo
\end{proof}

>>>>

\begin{definition}[Equivalent path graphs]
Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
\end{definition}

@@

Pruned de~Bruijn graphs are similar to manifold de Bruijn graphs \cite{Lin2014}.

\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P' = v'_{0} \dotsm v'_{\abs{P}-1}$ be a path in graph $G'$ with $1 \le \abs{P'} \le k$ and $G'.\glabel(P') \in (\Sigma \setminus \set{\#, \$})^{\ast}$. Then there is a path $P$ in $G$ with $G.\glabel(P) = G'.\glabel(P')$. If $G'$ is a de~Bruijn graph, $G'.\glabel(P)$ is a prefix of $G'.\gkey(v'_{0})$.
\end{lemma}

\begin{proof}
We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.

If $\abs{P} = 1$, every node $v \in G'.\gvalue(v'_{0})$ is a path in graph $G$ with label $X = G'.\glabel(P')$.
Otherwise every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$. Furthermore, the substring has $G'.\gkey(v'_{i})$ as its prefix. We can show by induction from $\abs{P}-2$ to $0$ that $S_{0}[j, j+k]$ has $X$ as its prefix. The sequence of nodes $P = \mathcal{S}.\gnode(S_{0}, j+i) \dotsm \mathcal{S}.\gnode(S_{0}, j+\abs{X}-1)$ is a path in graph $G$ with $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
\end{lemma}

\begin{proof}
Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs. Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.

Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$. By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring. We extend the substring to length $\abs{X}+k-1$ without changing its starting position. The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$. If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, there must be node $u \in V$ with $G.\glabel(u) = X[j]$ and with an edge $(u, v) \in E$ to a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$. Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key. As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$. When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.

Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$. We can also start from path $P''$ and determine the node $v_{0}$. As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
\end{proof}

De~Bruijn graphs are a special case of pruned de~Bruijn graphs. Hence we may compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
